Software Development in Startup Companies: The Green-field Startup Model

### 2013-03-04 article/main.tex
Revision 0.4. Sent to Tony for review

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
%\boldmath
Software startups are newly created companies with no operating history and
oriented in producing cutting-edge technologies. These companies develop
software under highly uncertain conditions, tackling fast growing markets with
severe lack of resources. Despite their increasing importance in economy there
are only few scientific studies attempting to address software engineering (SE)
issues, especially for early-stage startups. This research aims to understand
the software development strategies conducted by practitioners, in the period of
time that goes from idea conception to the first open beta release. A
\textit{grounded theory} approach guided the execution of a case study to
explore the state-of-practice. As result, the Greenfield startup model (GSM)
reveals the urgent priority of startups of releasing the product as quickly as
possible to verify the product/market fit and to adjust the business and product
trajectory according to early collected user feedback. Nevertheless, the need of
shortening time-to-market, by speeding-up the development through low-precision
and product-centric engineering activities, is counterbalanced by the need of
restructuring the product before setting off for further growth.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway. In particular, the Computer Society does
% not want either math or citations to appear in the abstract.

@@

From a software engineering perspective development in startups is challenging
since they develop software in a context where processes can hardly follow a
prescriptive methodology~\cite{Sutton2000, Coleman2005}. Despite startups share
some characteristics with similar domains (e.g. small and web companies), the
combination of different factors makes the specific development context quite
unique~\cite{Blank2005, Sutton2000}. Therefore, research is needed to support
their engineering activities~\cite{Coleman2005}, guiding practitioners in taking
decisions and avoiding choices that could easily lead the whole business to
failure~\cite{Kajko-Mattsson2008}. However, despite the impressive size of the
startup ecosystem~\cite{ISI:000243253000007}, the state-of-the-art presents a
gap. Through a \textit{Systematic Mapping Study} (SMS) of the literature,
conducted by Paternoster et al.~\cite{SMS}, only few publications were found
related to engineering activities in startups. Moreover the majority of these
studies do not possess the relevance and rigour required to form a consistent
body of knowledge in the state-of-the-art and in the state-of-practice.

This research aims to understand how software development strategies are
engineered by practitioners in startup companies in terms of level of:
structure, planning and control of software projects, in the period of time
that goes from idea conception to the first open beta release of the software
product.

With a cross-sectional study we focused the research context in a limited 
time-frame to emphasize the nature of uncertainty in the development activities 
and time pressure from the market, as discussed in~\cite{Maccormack2001,
Eisenhardt1998}. We performed semi-structured in-depth interviews with CEOs and
CTOs of startups, covering a wide spectrum of thematics and iteratively
adjusting the direction of the research according to the emerging evidences.

@@

The remainder of this paper is structured as follows: we provide relevant
information about startups in section~\ref{backg}. Section~\ref{resmet}
introduces the research questions and discusses the design and execution of the
research process. Results are discussed in section~\ref{res:gsm}, which presents
the GSM. Section~\ref{res:val} compares results of this research to literature
and frameworks presented in this field. Section~\ref{valt} discusses the
validity threats considered during this study. Section~\ref{sect:theory:impl}
discusses the most relevant implications of the GSM. Finally, this paper
concludes in section~\ref{conc}.

@@

Modern entrepreneurship has been boosted by the advent of the consumer internet
markets in the middle of the 1990s and culminated with the notorious \textit
{dot-com bubble burst} of 2000~\cite{Perkins:1999:IBI:555126,
storey1982entrepreneurship}. Several years later, with the massification of the
internet and mobile devices, we are now assisting to an impressive proliferation
of software ventures - metaphorically referred as the \textit{startup bubble}.
The easy access to vast potential markets and the low cost of services
distribution are appealing condition for modern entrepreneurs~\cite{Marmer2011}.
Inspired by success stories, a large number of software businesses are created
every day. However, the great majority of these companies fail within two years
from their creation~\cite{Crowne2002}.

@@

Moreover, most startups develop packaged applications rather than software for
a specific client~\cite{genome2012}. Issues related to this domain are addressed
in literature by the area known as \textit{market-driven software development}
(sometimes called \textit{packaged software development} or \textit{COTS
software development}~\cite{Alves2006}). Among other results, researchers
emphasize the importance of time-to-market as a key strategic 
objective~\cite{dagMDR,sawyer99} for companies operating in this domain. Other 
peculiar aspects which influence the software development are related to 
requirements, which are reported to be often ``invented by the software 
company''~\cite{512553}, ``rarely documented''~\cite{Karlsson02challengesin}, 
and can be validated only after the product is released to 
market~\cite{dahl2003,Keil:1995}. Under these circumstances, failure of product
launches are largely due to ``products not meeting customer 
needs''~\cite{Alves2006}.

@@

%------------------------------------------------------------------------------%
\subsection{General lack of research in startups} Despite some studies tried to
address the above mentioned issues, a SMS~\cite{SMS} revealed only a few SE
works in this specific area, as confirmed by other studies~\cite{Coleman2008,
Coleman2008a, Coleman2007, Sutton2000}. Moreover, the studies, identified 
in~\cite{SMS}, appear to be highly fragmented and spread across different areas
rather than constituting a consistent \textit{body of knowledge}.

A research published in 2000 by Sutton~\cite{Sutton2000} confirmed a general
lack of studies in this area, claiming that \textit{``software startups
represent a segment that has been mostly neglected in process studies''} and it
has been further confirmed with the empirical studies of Coleman et 
al.~\cite{Coleman2008,Coleman2008a,Coleman2007} eight years later.

@@

A study of Kajko-Mattsson~\cite{Kajko-Mattsson2008}, which investigated a
Swedish software startup, reported a heavy lack of requirements gathering
processes, minimal project management, lack of control over the change requests,
absence of documentation to track the status and progress of the process, and
defective releases. Accordingly, Ambler et al. report how two startups
approaching to an upcoming IPO started to require processes to focus on scalable
solutions, in view of the growing company's size in terms of users and 
employees~\cite{Ambler2002}. In this regard, Crowne~\cite{Crowne2002} specifies 
different stages through which software startups evolve. Starting without any 
established workflows, startups grow over time, creating and stabilizing 
processes to eventually improving them only when sufficiently mature.

But when startups have no time for training and orienting activities, as
discussed in~\cite{Sutton2000}, their main focus remains on team capabilities
instead of prescriptive processes, hiring people who can \textit{``hit the
ground running''}~\cite{Yoffie1999}. Empowering the team and focus on
methodological attributes of the processes oriented in prototyping, 
proof-of-concepts, mock-ups and demos, to test basic functionalities, have been  
the primary priority in startups~\cite{Camel1994a}. Only when they grow, formal
methodologies arise, followed by performing quality assurance and long-term
planning processes~\cite{Yoffie1999}.

@@

As partially discussed above, contributions to flexibility and reactiveness of
the development process has been conducted prominently by means of 
\textit{Lean}~\cite{Gautam2008} and \textit{Agile}~\cite{Abrahamsson2002} 
methodologies (also reported in~\cite{Taipale2010, Kuvinka2011}), where the 
extreme uncertain conditions lead startups to learn fast from trials and errors 
with a strong customer relationship in order to avoid wasting time in building 
wrong functionalities and prevent rapidly exhaustion of 
resources~\cite{Midler2008,Hilmola2003, Sutton2000}. Customer involvement in 
software development has also been discussed by Yogendra~\cite{Yogendra2002} as 
an important factor to encourage an early alignment of business concerns to the 
technology strategies, because both are salient considerations to be 
successful~\cite{Tingling2007}.

@@

When startups are mature enough to support software process improvement (SPI),
the solutions considered according to the state-of-the-art are oriented towards
light-weight processes such as a design of development process based on XP,
proposed by Zettel~\cite{Zettel2001}. The process consists of a set of
activities and artifacts (in addition to some important roles) defined in order
to identify responsibilities and tools to utilize. But, despite the promising
benefits reported by Zettel, we were not able to identify any future evaluation
in real-world settings.

Another attempt of SPI in startups has been conducted by Deakins et al.
introducing a \textit{Helical Model for managing e-commerce development
environment}~\cite{Deakins2005}. Also in this case, the author prescribed broad
guidelines for a rapid high-quality development process, which underwent limited
testing only in academic settings.

The problem of \textit{one-size-fits-all}, related to the SPI representations
for startups, is described by Fayad~\cite{Fayad1997}. The author discusses the
problem in actuating the same \textit{best-practices} criteria for established
companies in 10-person software startups. Thoroughly remarked 
in~\cite{Sutton2000}, Sutton states that problems of SPI in software startups 
arise because of: the dynamism of the development process, which precludes
repeatability; organizational maturity, that cannot be maintained from startups
in view of lack of corporate direction; severe lack of resources, both human and
technological for process definition, implementation, management, and training.
In conclusion, the primary benefits of SPI do not address startups, which
instead of promoting product quality, aim to minimize time-to-market.

Additionally, the role of SPI has been neglected because seen as an obstacle to
the development team's creativity and flexibility, and to the need of a quick
delivering product process environment~\cite{Coleman2008a}. 
%MUN: the second part of the sentence is just a statement without reference. If 
%it refers to a statement by Coleman, put his reference at the end OR find the  
%correct reference OR delete the statement. %CG fixed 
Product quality is left aside in favor of minimal and suitable functionalities 
to shorten the time-to-market. As reported in two studies of Mater and 
Mirel~\cite{Mater2000, Mirel2000}, quality aspects, mostly taken in 
consideration in internet startups, are oriented to usability and scalability, 
even though the market and application type heavily influences the 
quality-demand~\cite{Coleman2008, Kim2005}.

@@

%------------------------------------------------------------------------------%
\subsection{Technical debt} Notwithstanding, a new area of SE research,
trying to tackle the problem of the \textit{technical debt}, brings and
encompasses different implications in studying development in software 
startups~\cite{Tom2013}. The metaphoric neologism was originally introduced by
Cunningham in 1992~\cite{TechnicalDebtCunn} and has recently attracted the
attention of SE researchers\footnote{To attest the fact that \textit{technical
debt} is gaining traction among researchers, we mention two important
contributions which characterize the \textit{``debt landscape''}:
\cite{Nugroho2011,Izurieta2012} and a dedicated workshop~\cite{workshopDebt}
organized by the Software Engineering Institute and ICSE.}. The concept of
\textit{technical debt} is well illustrated 
in~\cite{Brown:2010:MTD:1882362.1882373}: \textit{``The idea is that developers
sometimes accept compromises in a system in one dimension (e.g., modularity) to
meet an urgent demand in some other dimension (e.g., a deadline), and that such
compromises incur a ``debt'' on which ``interest'' has to be paid and which the
``principal'' should be repaid at some point for the long-term health of the
project''}. The compromise between high-speed and high-quality engineering is
faced daily by software startups, not only in terms of architecture design but
in multifaceted aspects (weak project management, testing, process control).

@@

\begin{compactitem} 
\item Software development strategy: the overall approach adopted by the 
company to carry out the product development. 
\item Engineering activities: the activities needed to bring a product from 
idea to market. Traditional engineering activities are, among others, 
requirement engineering, design, architecture, implementation, testing. 
\item Engineering elements: any practice, tool and artifacts contributing and 
supporting the engineering activities.
\item Quality attributes: those overall factors that affect run-time
behavior, system design, and user experience. They represent areas of concern
that have the potential for applications to impact across layers and tiers.
Some of these attributes are related to the overall system design, while others
are specific to run time, design time, or user centric 
issues~\cite{Microsoft2009}. 
\item Growth: an increase in the company size respect to the initial 
conditions in terms of either employees or users/customers, and
product complexity in terms of handling an increasing number of feature
requests. 
\item Software product: any software product and software service.
\end{compactitem}

@@

The aim of this research is to \textit{understand how software development
strategies are engineered by practitioners in startup companies in terms of
level of: structure, planning and control of software projects, in the period of
time that goes from idea conception to the first open beta release of the
software product}\footnote{The above mentioned goal is structured according to
the GQM paradigm~\cite{Basili1992}.}.

Initially the boundaries of the research domain were set by means of 
non-systematic literature survey, which provided the initial keywords used  
further in the research process. Moreover it helped us in defining the research
questions (RQs):

@@

\textit{Grounded Theory} methodology is described by Robson as \textit{``the
best approach to answer the question - What is going on here?''} and defined by
its creators~\cite{ColinRobson2009} as \textit{``a set of well-developed
categories ( e.g. themes, concepts ) that are systematically interrelated
through statements of relations to form a theoretical framework that explains
some relevant social, psychological, educational, nursing or other
phenomenon''}.

@@

Following the GT principles we attempted to capture the most relevant aspects
of software development from startup practitioners, letting a theory emerge
from the interviews and adjusting the research hypotheses and questions as we
proceeded through. During these interviews we collected data related to
engineering activities undertaken by startups in the time frame between the idea
conception and the first open beta release of the product. Then, we proceeded
to the analysis of the data, finding important relations among concepts with a
formal approach.

As suggested by Coleman, in view of the different versions of GT which emerged
in the last years, researchers should indicate which \textit{``implementation''}
of the theory is being used~\cite{Coleman2007}. After the methodology was
initially introduced, its original authors had divergences regarding how the
theory should be formed by analyzing the data (Strauss and Corbin on one side,
Glaser on the other). Glaser advocated for a more \textit{puristic} approach,
where the theoretical categories should \textit{``naturally emerge from the
data''}, whilst Strauss and Corbin formulated an updated version of the
methodology, which leaves to the researchers an higher degree of freedom. The
latter approach empowers researchers' \textit{``theoretical 
sensitivity''}~\cite{Corbin1990}, and encourages them to outline the research 
problem beforehand. Since the knowledge obtained from parallel execution of the 
SMS~\cite{SMS} and our direct experience with startups companies provided a good
initial level of knowledge, in this study we use the Corbin and Strauss
approach~\cite{Strauss1998}.

@@

The whole procedure has been executed simultaneously in pair on the same screen
by the first two authors, handling conflicts by reviewing the rationale of
decision with the third and fourth authors. When necessary we performed an in-
depth review of the design of the research and data collected during the
execution process.

@@

This case study started based on a SMS of the literature regarding software
development in startups~\cite{SMS}. By means of this SMS we refined our research
questions. The research questions have been narrowed down to outline the domain
of our study, even though still broad enough to allow emerging new theories and
to adjust the scope of the research~\cite{Corbin1990}.

@@

All companies, except C10, were founded within the last three years (2009-2012),
%MUN: I've shortened the text and added absolute numbers. Is 2009-2012 correct?
%CG: right by an average of 3 starting members, which were in majority
developers. Moreover, the number of current employees shows how, to different
degrees, companies expanded the initial teams. All companies, except C5,
brought their first product to market in less than 6 months from the idea
conception. The products consist of web-applications, launched in different
nations and market sectors. The growing team size and the publicly available
data suggest an healthy status of the businesses.

@@

%------------------------------------------------------------------------------%
\subsection{Data collection} The data collection was designed to allow
triangulation, which integrates multiple data sources converging on the same
phenomenon. After transcribing the interview, we extracted conceptual
relations, integrating them with the questionnaire's results. A well structured
case-study database allowed us to easily retrieve and seek for information,
assembling the evidence from the case study reports, as described also 
in~\cite{Yin1994}. The database has been stored and constructed using the
qualitative data analysis software package AtlasTI\footnote{Available online at
\url{http://www.atlasti.com/}.} - one of the most suitable tools for 
GT~\cite{Coleman2007}. We overlapped interviews with questionnaire results to
adjust the data collection and take advantage of emergent themes and reveal
possible inconsistencies. The data was analyzed simultaneously and with
flexibility in mind in such a way that adjustments were made according to the
emerging findings.
%------------------------------------------------------------------------------%
\subsection{Data analysis} Before starting the analysis, a data ordering
procedure was necessary since interviews were spread across a multitude of
topics. Therefore transcripts have been structured in thematic areas
accordingly to different topic cards used during the interviews. We proceeded
horizontally between same thematic areas of transcripts to be able to identify
a better number of similar concepts, rather than going through an entire
transcript at one time.

@@

As we were iterating through the interviews, we analyzed new data accordingly,
by updating codes and categories on necessity, and taking notes in the form of
memos to adjust the new emerging theory. To improve the speed of the coding
process, we analyzed all the transcripts one part at-the-time: transcripts
followed the semi-structured interview format, and this approach enhanced the
chances of finding similar codes across different transcripts, allowing us to
quickly iterate and update categories.

@@

Strauss and Corbin provided a list of questions to assist in determining how
well the findings are grounded~\cite{Strauss1998}. They questions are as
follows:

\par\vspace {0.4cm} 
\begin{compactenum} 
\item Are concepts generated, and are the concepts systematically related? 
\item Are there many conceptual linkages and are the categories well developed? 
\item Is variation\footnote{Variation refers to the variety of contexts to 
which the theory can be applied.} built into the theory and are the conditions 
under which variation can be found built into the study and explained? 
\item Are the conditions under which variation can be found built into the 
study and explained? 
\item Has the process been taken into account, and does the theory stand the 
test of time? 
\item Do the theoretical findings seem significant, and to what extent? 
\end{compactenum}
\par\vspace {0.4cm}

Answering the above questions, the concepts were generated according to the
coding process. They were systematically related through the use of a network
diagram (see Figure~\ref{fig:gt:completemethodology}), by which we established
the relations between the concepts. At each iteration of the grounded theory
process, we considered and examined a concept within different conditions and
dimensions, trying to incorporate data from a broader range of practitioners.

@@

In the following subsections we explain in detail all the categories presented
in Figure~\ref{fig:gsm}, and conclude in subsection~\ref{res:gsm:th} with the
final theory. In the explanations we use identifiers of the companies presented
in Table~\ref{t_interviews-stats} (i.e. Cx) to highlight statements recorded
from the interviewees.

@@

To validate the generalization of the theory, we describe conceptualizations
defined from the GSM that are in support of previous models developed by
Coleman~\cite{Coleman2007,Coleman2008a, Coleman2008}, 
Baskerville~\cite{Internet} and finally Brooks~\cite{BrooksJr1987}.

>>>>

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
%\boldmath
Software startups are newly created companies with no operating history and
oriented in producing cutting-edge technologies. These companies develop
software under highly uncertain conditions, tackling fast growing markets with
severe lack of resources. Despite their increasing importance in economy there
are only few scientific studies attempting to address software engineering (SE)
issues, especially for early-stage startups. This research aims to understand
the software development strategies conducted by practitioners, in the period of
time that goes from idea conception to the first open beta release. A grounded 
theory approach guided the execution of a case study to explore the 
state-of-practice. As result, the Greenfield startup model (GSM) reveals the 
urgent priority of startups of releasing the product as quickly as possible to 
verify the product/market fit and to adjust the business and product trajectory 
according to early collected user feedback. Nevertheless, the need of 
shortening time-to-market, by speeding-up the development through low-precision
and product-centric engineering activities, is counterbalanced by the need of
restructuring the product before setting off for further growth.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway. In particular, the Computer Society does
% not want either math or citations to appear in the abstract.

@@

From a software engineering perspective, development in startups is challenging
since they develop software in a context where processes can hardly follow a
prescriptive methodology~\cite{Sutton2000, Coleman2005}. Despite startups share
some characteristics with similar domains (e.g. small and web companies), the
combination of different factors makes the specific development context quite
unique~\cite{Blank2005,Sutton2000}. Therefore, research is needed to support
their engineering activities~\cite{Coleman2005}, guiding practitioners in taking
decisions and avoiding choices that could easily lead the whole business to
failure~\cite{Kajko-Mattsson2008}. However, despite the impressive size of the
startup ecosystem~\cite{ISI:000243253000007}, the state-of-the-art presents a
gap. Through a \textit{Systematic Mapping Study} (SMS) of the literature,
conducted by Paternoster et al.~\cite{SMS}, only few publications were found
related to engineering activities in startups. Moreover the majority of these
studies do not possess the relevance and rigour required to form a consistent
body of knowledge in the state-of-the-art and in the state-of-practice.

This research aims to understand how software development strategies are
engineered by practitioners in startup companies in terms of: structure, 
planning and control of software projects, in the period of time that goes from 
idea conception to the first open beta release of the software product.

With a cross-sectional study we focused the research context on a limited 
time-frame to emphasize the nature of uncertainty in the development activities 
and time pressure from the market, as discussed in~\cite{Maccormack2001,
Eisenhardt1998}. We performed semi-structured, in-depth interviews with CEOs and
CTOs of startups, covering a wide spectrum of themes and iteratively
adjusting the direction of the research according to the emerging evidence.

@@

The remainder of this paper is structured as follows: we provide relevant
information about startups in section~\ref{backg}. Section~\ref{resmet}
introduces the research questions and discusses the design and execution of the
research process. Results are discussed in section~\ref{res:gsm}, which presents
the GSM. Section~\ref{res:val} compares results of this research to literature
and frameworks presented in this field. Section~\ref{sect:theory:impl}
discusses the most relevant implications of the GSM. Section~\ref{valt} 
discusses the validity threats considered during this study. The paper
concludes in section~\ref{conc}.

@@

Modern entrepreneurship has been boosted by the advent of the consumer internet
markets in the middle of the 1990s and culminated with the notorious 
\textit{dot-com bubble burst} of 2000~\cite{Perkins:1999:IBI:555126,
storey1982entrepreneurship}. Several years later, with the massification of the
internet and mobile devices, we are now assisting to an impressive proliferation
of software ventures - metaphorically referred to as the \textit{startup 
bubble}. The easy access to vast potential markets and the low cost of services
distribution are appealing conditions for modern
entrepreneurs~\cite{Marmer2011}. Inspired by success stories, a large number of 
software businesses are created every day. However, the great majority of these 
companies fail within two years from their creation~\cite{Crowne2002}.

@@

Moreover, most startups develop packaged applications rather than software for
a specific client~\cite{genome2012}. Issues related to this domain are addressed
in literature by \textit{market-driven software development}
(sometimes called \textit{packaged software development} or \textit{COTS
software development}~\cite{Alves2006}). Among other results, researchers
emphasize the importance of time-to-market as a key strategic 
objective~\cite{dagMDR,sawyer99} for companies operating in this domain. Other 
peculiar aspects which influence the software development are related to 
requirements, which are reported to be often ``invented by the software 
company''~\cite{512553}, ``rarely documented''~\cite{Karlsson02challengesin}, 
and can be validated only after the product is released to 
market~\cite{dahl2003,Keil:1995}. Under these circumstances, failure of product
launches are largely due to ``products not meeting customer 
needs''~\cite{Alves2006}.

@@

%------------------------------------------------------------------------------%
\subsection{General lack of research in startups} Despite some studies tried to
address the above mentioned issues, a SMS~\cite{SMS} revealed only a few SE
works in this specific area, as confirmed by other studies~\cite{Coleman2008,
Coleman2008a, Coleman2007, Sutton2000}. Moreover, the studies appear to be 
highly fragmented and spread across different areas rather than constituting a 
consistent \textit{body of knowledge}.

Sutton~\cite{Sutton2000} confirmed in 2000 a general lack of studies in this 
area, claiming that \textit{``software startups represent a segment that has 
been mostly neglected in process studies''} and it has been further confirmed 
with the empirical studies by Coleman et 
al.~\cite{Coleman2008,Coleman2008a,Coleman2007} eight years later.

@@

A study by Kajko-Mattsson~\cite{Kajko-Mattsson2008}, investigating a
Swedish software startup, reported a heavy lack of requirements gathering
processes, minimal project management, lack of control over the change requests,
absence of documentation to track the status and progress of the process, and
defective releases. Accordingly, Ambler et al.~\cite{Ambler2002} reported how 
two startups approaching to an upcoming IPO started to require processes to 
focus on scalable solutions, in view of the growing company's size in terms of 
users and employees. In this regard, Crowne~\cite{Crowne2002} specifies 
different stages through which software startups evolve. Starting without any 
established workflows, startups grow over time, creating and stabilizing 
processes to eventually improving them only when sufficiently mature.

But when startups have no time for training and orienting activities, as
discussed by Sutton~\cite{Sutton2000}, their main focus remains on team 
capabilities instead of prescriptive processes, hiring people who can 
\textit{``hit the ground running''}~\cite{Yoffie1999}. Empowering the team and 
focus on methodological attributes of the processes oriented towards 
prototyping, proof-of-concepts, mock-ups and demos, to test basic 
functionalities, have been  the primary priority in startups~\cite{Camel1994a}. 
Only when they grow, formal
methodologies arise, followed by performing quality assurance and long-term
planning processes~\cite{Yoffie1999}.

@@

Contributions to flexibility and reactiveness of the development process has 
been conducted prominently by means of \textit{Lean}~\cite{Gautam2008} and 
\textit{Agile}~\cite{Abrahamsson2002} methodologies (also reported 
in~\cite{Taipale2010, Kuvinka2011}), where the extreme uncertain conditions lead 
startups to learn fast from trial and error with a strong customer 
relationship in order to avoid wasting time in building wrong functionalities 
and prevent rapidly exhaustion of resources~\cite{Midler2008,Hilmola2003, 
Sutton2000}. Customer involvement in software development has also been 
discussed by Yogendra~\cite{Yogendra2002} as an important factor to encourage an 
early alignment of business concerns to the technology strategies, because both 
are salient considerations to be successful~\cite{Tingling2007}.

@@

When startups are mature enough to support software process improvement (SPI),
the solutions considered according to the state-of-the-art are oriented towards
light-weight processes such as a design of development process based on XP,
proposed by Zettel~\cite{Zettel2001}. The process consists of a set of
activities and artifacts to identify responsibilities and tools to utilize. But, 
despite the promising benefits reported by Zettel, we were not able to identify 
any evaluation in real-world settings. Another attempt of SPI in startups has 
been conducted by Deakins et al.~\cite{Deakins2005}, introducing broad
guidelines for a rapid high-quality development process, which underwent 
however only limited testing in academic settings.

The problem of \textit{one-size-fits-all}, related to the SPI representations
for startups, is described by Fayad~\cite{Fayad1997}. The author discusses the
problem in actuating the same best-practices criteria for established
companies in 10-person software startups. Sutton~\cite{Sutton2000} remarks that 
problems of SPI in software startups arise due to: the dynamism of the 
development process, which precludes repeatability; organizational maturity, 
that cannot be maintained from startups in view of lack of corporate direction; 
severe lack of resources, both human and technological for process definition, 
implementation, management, and training. In conclusion, the primary benefits 
of SPI do not address startups, which instead of promoting product quality, aim 
to minimize time-to-market.

Additionally, the role of SPI has been neglected because it is seen as an 
obstacle to the development team's creativity and flexibility, and to the need 
of a quick delivering product process environment~\cite{Coleman2008a}. 
%MUN: the second part of the sentence is just a statement without reference. If 
%it refers to a statement by Coleman, put his reference at the end OR find the  
%correct reference OR delete the statement. %CG fixed 
Product quality is left aside in favor of minimal and suitable functionalities 
to shorten the time-to-market. As reported in two studies by Mater and 
Mirel~\cite{Mater2000, Mirel2000}, quality aspects mostly taken in 
consideration in internet startups, are oriented towards usability and 
scalability, even though the market and application type heavily influences the 
quality-demand~\cite{Coleman2008, Kim2005}.

@@

%------------------------------------------------------------------------------%
\subsection{Technical debt} 
Notwithstanding, a new area of SE research, trying to tackle the problem of 
\textit{technical debt}, brings and encompasses different implications in 
studying development in software startups~\cite{Tom2013}. The metaphoric 
neologism was originally introduced by Cunningham in 
1992~\cite{TechnicalDebtCunn} and has recently attracted the
attention of SE researchers\footnote{To attest that \textit{technical
debt} is gaining traction among researchers, we mention two important
contributions which characterize the \textit{``debt landscape''}:
\cite{Nugroho2011,Izurieta2012} and a dedicated workshop~\cite{workshopDebt}
organized by the Software Engineering Institute and ICSE.}. The concept of
technical debt is well illustrated in~\cite{Brown:2010:MTD:1882362.1882373}: 
\textit{``The idea is that developers sometimes accept compromises in a system 
in one dimension (e.g., modularity) to meet an urgent demand in some other 
dimension (e.g., a deadline), and that such compromises incur a ``debt'' on 
which ``interest'' has to be paid and which the ``principal'' should be repaid 
at some point for the long-term health of the project''}. The compromise 
between high-speed and high-quality engineering is faced daily by software 
startups, not only in terms of architecture design but in multifaceted aspects 
(weak project management, testing, process control).

@@

\begin{compactitem} 
\item Software development strategy: the overall approach adopted by the 
company to carry out the product development. 
\item Engineering activities: the activities needed to bring a product from 
idea to market. Traditional engineering activities are, among others, 
requirement engineering, design, architecture, implementation, testing. 
\item Engineering elements: any practice, tool and artifacts contributing and 
supporting the engineering activities.
\item Quality attributes: those overall factors that affect run-time
behavior, system design, and user experience. They represent areas of concern
that have the potential for applications to impact across layers and tiers.
Some of these attributes are related to the overall system design, while others
are specific to run time, design time, or user centric 
issues~\cite{Microsoft2009}. 
\item Growth: an increase in the company size with respect to the initial 
conditions in terms of either employees or users/customers, and
product complexity in terms of handling an increasing number of feature
requests. 
\item Software product: any software product and/or software service.
\end{compactitem}

@@

The aim of this research is to \textit{understand how software development
strategies are engineered by practitioners in startup companies in terms of: 
structure, planning and control of software projects, in the period of
time that goes from idea conception to the first open beta release of the
software product}\footnote{The above mentioned goal is structured according to
the GQM paradigm~\cite{Basili1992}.}.

Initially, the boundaries of the research domain were set by means of a
non-systematic literature survey, which provided the initial keywords used  
further in the research process. Moreover it helped us in defining the research
questions (RQs):

@@

\textit{Grounded Theory} methodology is described by Robson as \textit{``the
best approach to answer the question - What is going on here?''} and defined by
its creators~\cite{ColinRobson2009} as \textit{``a set of well-developed
categories (e.g. themes, concepts) that are systematically interrelated
through statements of relations to form a theoretical framework that explains
some relevant social, psychological, educational, nursing or other
phenomenon''}.

@@

Following the GT principles, we captured the most relevant aspects of software 
development from startup practitioners, letting a theory emerge from the 
interviews and adjusting the research hypotheses and questions as we
proceeded. During these interviews we collected data related to
engineering activities undertaken by startups in the time frame between the idea
conception and the first open beta release of the product. Then, we proceeded
to the analysis of the data, finding important relations among concepts with a
formal approach.

As suggested by Coleman, in view of the different versions of GT, researchers 
should indicate which \textit{``implementation''} of the theory is being 
used~\cite{Coleman2007}. After the GT methodology was initially introduced, its 
original authors had divergences regarding how the theory should be formed by 
analyzing the data (Strauss and Corbin on one side, Glaser on the other). 
Glaser advocated for a more \textit{puristic} approach, where the theoretical 
categories should \textit{``naturally emerge from the data''}, whilst Strauss 
and Corbin formulated an updated version of the methodology, which leaves to 
the researchers an higher degree of freedom. The latter approach empowers 
researchers' \textit{``theoretical sensitivity''}~\cite{Corbin1990}, and 
encourages them to outline the research problem beforehand. Since the knowledge 
obtained from parallel execution of the SMS~\cite{SMS} and our direct experience 
with startups companies provided a good
initial level of knowledge, in this study we use the Corbin and Strauss
approach~\cite{Strauss1998}.

@@

The whole procedure has been executed simultaneously in pair on the same screen
by the first two authors, handling conflicts by reviewing the rationale of
decision with the third and fourth authors. When necessary we performed an 
in-depth review of the design of the research and data collected during the
execution process.

@@

This case study started based on a SMS of the literature regarding software
development in startups~\cite{SMS}, refining our research questions. The 
research questions have been narrowed down to outline the domain
of our study, even though still broad enough to allow new theories to emerge and
to adjust the scope of the research~\cite{Corbin1990}.

@@

All companies, except C10, were founded within the last three years (2009-2012),
%MUN: I've shortened the text and added absolute numbers. Is 2009-2012 correct?
%CG: right,
by an average of 3 starting members, which were in majority developers. 
Moreover, the number of current employees shows how, to different
degrees, companies expanded the initial teams. All companies, except C5,
brought their first product to market within less than 6 months from the idea
conception. The products consist of web-applications, launched in different
nations and market sectors. The growing team size and the publicly available
data suggest an healthy status of the businesses.

@@

%------------------------------------------------------------------------------%
\subsection{Data collection} 
The data collection was designed to allow triangulation, which integrates 
multiple data sources converging on the same phenomenon. After transcribing the 
interview, we extracted conceptual relations, integrating them with the 
questionnaire's results. A well structured case-study database allowed us to 
easily retrieve and seek for information, assembling the evidence from the case 
study reports, as described also by Yin~\cite{Yin1994}. The database has been 
stored and constructed using the qualitative data analysis software package 
AtlasTI\footnote{Available online at \url{http://www.atlasti.com/}.} - one of 
the most suitable tools for GT~\cite{Coleman2007}. We overlapped interviews with 
questionnaire results to adjust the data collection and take advantage of 
emergent themes and reveal possible inconsistencies. The data was analyzed 
simultaneously and with flexibility in mind in such a way that adjustments were 
made according to the emerging findings.

%------------------------------------------------------------------------------%
\subsection{Data analysis} 
Before starting the analysis, a data ordering procedure was necessary since 
interviews were spread across a multitude of topics. Therefore transcripts have 
been structured in thematic areas accordingly to different topic cards used 
during the interviews. We proceeded horizontally between same thematic areas of 
transcripts to be able to identify a better number of similar concepts, rather 
than going through an entire transcript at one time.

@@

As we were iterating through the interviews, we analyzed new data accordingly,
by updating codes and categories on necessity, and taking notes in the form of
memos to adjust the new emerging theory. To improve the speed of the coding
process, we analyzed all the transcripts one part at-a-time: transcripts
followed the semi-structured interview format, and this approach enhanced the
chances of finding similar codes across different transcripts, allowing us to
quickly iterate and update categories.

@@

Strauss and Corbin provided a list of questions to assist in determining how
well the findings are grounded~\cite{Strauss1998}:

\begin{compactenum} 
\item Are concepts generated, and are the concepts systematically related? 
\item Are there many conceptual linkages and are the categories well developed? 
\item Is variation\footnote{Variation refers to the variety of contexts to 
which the theory can be applied.} built into the theory and are the conditions 
under which variation can be found built into the study and explained? 
\item Are the conditions under which variation can be found built into the 
study and explained? 
\item Has the process been taken into account, and does the theory stand the 
test of time? 
\item Do the theoretical findings seem significant, and to what extent? 
\end{compactenum}

Answering the above questions, the concepts were generated according to the
coding process. They were systematically related through the use of a network
diagram by which we established the relations between the concepts. At each 
iteration of the grounded theory process, we considered and examined a concept 
within different conditions and dimensions, trying to incorporate data from a 
broader range of practitioners.

@@

In the following subsections we explain the categories presented in 
Figure~\ref{fig:gsm}, and conclude in subsection~\ref{res:gsm:th} with the
final theory. In the explanations we use identifiers of the companies presented
in Table~\ref{t_interviews-stats} (i.e. Cx) to highlight statements recorded
from the interviewees.

@@

To validate the generalization of the theory, we describe conceptualizations
defined from the GSM that are in support of previous models developed by
Coleman~\cite{Coleman2007,Coleman2008a, Coleman2008}, 
Baskerville~\cite{Internet} and Brooks~\cite{BrooksJr1987}.
